---
title: "TRGN510_FinalProject"
output: html_document
---

## Find and set working directory
```{r}
getwd()
setwd("/Users/wendy/Desktop/finalproject")
```

## Load required packages
```{r}
library(Glimma)
library(edgeR)
library(limma)
library(Homo.sapiens)
```

## Load in the data
Black male and white male data are in different folders, so I make a 2D-list. I put all the file names in the black male folder in the first row of 2D-list, and all the file names in the white male folder in the second row of 2D-list. Then, I make a for loop to read each file in the folders.
```{r}
path <- paste("/Users/wendy/Desktop/final_project",sep="/")
folders <- list.dirs(path)
folders <- folders[(length(folders)-1):length(folders)]
store <- list()
for(i in 1:length(folders)){
  store = cbind(store, list.files(folders[i],pattern = "htseq.counts"))
}
```

## Group the data
Read each of the files into R and combine into a dataframe called as "x"
```{r}
x <- readDGE(store, columns=c(1,2))
class(x)
```

This funtion shows the dimensions of the dataframe. There are 60487 genes in each file and there are 30 files (15 black males and 15 white males).
```{r}
dim(x)
```

## Organize sample information
DGEList-oject contains a `samples` dataframe. The samplenames are complicated, so I name them black male 1 to 15, and white male 1 to 15.
```{r}
samplenames <- c(rep("black_male",15), rep("white_male",15))
for( i in 1:(length(samplenames)/2)){ 
  samplenames[i] = paste(samplenames[i], as.character(i), sep='_')
}
for(i in 1:(length(samplenames)/2)){
  samplenames[length(samplenames)/2+i] = paste(samplenames[length(samplenames)/2+i], as.character(i), sep='_')
}
colnames(x) <- samplenames
```

The samples are grouped according to their ethnicity.
```{r}
group <- c(rep("black_male",15), rep("white_male",15))
x$samples$group <- group
x$samples
```

## Organize gene annotations
Remove the all the digits after the decimal of Ensembl name, so it looks more similar to the key provided.
```{r}
geneid <- rownames(x)
geneid <-gsub("\\.\\d+$", "", geneid)
head(geneid)
```

I load "Homo.sapiens" library to get the information for this dataframe. The GeneIDs are related to the gene symbol and chromosome number.
```{r}
genes <- select(Homo.sapiens, keys=geneid, columns=c("SYMBOL", "TXCHROM"), keytype="ENSEMBL")
head(genes)
```

Remove the duplicated genes from the dataset.
```{r}
genes <- genes[!duplicated(genes$ENSEMBL),]
x$genes <- genes
x
```

