---
title: "TRGN510_FinalProject"
output: html_document
---

## Find and set working directory
```{r}
getwd()
setwd("/Users/wendy/Desktop/TRGN510_FinalProject")
```

## Load required packages
```{r}
library(Glimma)
library(edgeR)
library(limma)
library(Homo.sapiens)
```

## Load in the data
Black male and white male data are in different folders, so I make a 2D-list. I put all the file names in the black male folder in the first row of 2D-list, and all the file names in the white male folder in the second row of 2D-list. Then, I make a for loop to read each file in the folders.
```{r}
path <- paste("/Users/wendy/Desktop/TRGN510_FinalProject",'data',sep="/")

folders <- list.dirs(path)
folders <- folders[2:length(folders)]
store <- list()
for(i in 1:length(folders)){
  store = cbind(store, list.files(folders[i],pattern = "htseq.counts"))
}
for(i in 1:length(folders)){
  store[,i] = paste(folders[i],store[,i],sep = '/')
}
```

## Group the data
Read each of the files into R and combine into a dataframe called as "x"
```{r}
x <- readDGE(store, columns=c(1,2))
class(x)
```

This funtion shows the dimensions of the dataframe. There are 60487 genes in each file and there are 30 files (15 black males and 15 white males).
```{r}
dim(x)
```

## Organize sample information
DGEList-oject contains a `samples` dataframe. The samplenames are complicated, so I name them black male 1 to 15, and white male 1 to 15.
```{r}
samplenames <- c(rep("black_male",15), rep("white_male",15))
for( i in 1:(length(samplenames)/2)){ 
  samplenames[i] = paste(samplenames[i], as.character(i), sep='_')
}
for(i in 1:(length(samplenames)/2)){
  samplenames[length(samplenames)/2+i] = paste(samplenames[length(samplenames)/2+i], as.character(i), sep='_')
}
sampleNames
colnames(x) <- samplenames
```

The samples are grouped according to their ethnicity.
```{r}
group <- c(rep("black_male",15), rep("white_male",15))
x$samples$group <- group
x$samples
```

## Organize gene annotations
I removed the all the digits after the decimal of Ensembl name, so it looks more similar to the key provided.
```{r}
geneid <- rownames(x)
geneid <-gsub("\\.\\d+$", "", geneid)
head(geneid)
```

I loaded "Homo.sapiens" library to get the information for this dataframe. The GeneIDs are related to the gene symbol and chromosome number.
```{r}
genes <- select(Homo.sapiens, keys=geneid, columns=c("SYMBOL", "TXCHROM"), keytype="ENSEMBL")
head(genes)
```

I removed the duplicated genes from the dataset.
```{r}
genes <- genes[!duplicated(genes$ENSEMBL),]
x$genes <- genes
x
```

## Data pre-processing
## Transformations from the raw-scale
I transformed the data from its raw form to a scale that accounts for such library size differences. "cpm" means counts per million, and "lcpm" means log of counts per million.
```{r}
cpm <- cpm(x)
lcpm <- cpm(x, log=TRUE)
```

L is the average library size in millions, and M is the median of the library size in millions. For this dataset, the average library size is about 87.8 million.
```{r}
L <- mean(x$samples$lib.size) * 1e-6
M <- median(x$samples$lib.size) * 1e-6
c(L, M)
```

This is the summary of the lcpm calculation.
```{r}
summary(lcpm)
```

## Removing genes that are lowly expressed
All datasets include genes that are expressed and not expressed, so we want to remove some genes that are lowly expressed and do not contribute to the statistical analysis.
```{r}
table(rowSums(x$counts==0)==9)
```

EdgeR package provides an easy way to remove these lowly expressed genes.
```{r}
keep.exprs <- filterByExpr(x, group=group)
x <- x[keep.exprs,, keep.lib.sizes=FALSE]
dim(x)
```

The number of genes is reduced from 60487 to 20902. The `RColorBrewer` package allows us to create the figure. The following code is to produce the figure that is raw data and filtered data.
```{r}
lcpm.cutoff <- log2(10/M + 2/L)
library(RColorBrewer)
nsamples <- ncol(x)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,2))
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
lcpm <- cpm(x, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
```

## Normalizing gene expression distributions
EdgeR package provides `calcNormFactors` function for statistical analysis. The normalization factors are stored in a dataset called `norm.factors`.
```{r}
x <- calcNormFactors(x, method = "TMM")
x$samples$norm.factors
```

```{r}
x2 <- x
x2$samples$norm.factors <- 1
x2$counts[,1] <- ceiling(x2$counts[,1]*0.05)
x2$counts[,2] <- x2$counts[,2]*5
```

The plot below shows the expression distribution of samples for unnormalized data.
```{r}
par(mfrow=c(1,2))
lcpm <- cpm(x2, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="A. Example: Unnormalised data",ylab="Log-cpm")
x2 <- calcNormFactors(x2)  
x2$samples$norm.factors
```

The plot below shows the expression distribution of samples for normalized data.
```{r}
lcpm <- cpm(x2, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="B. Example: Normalised data",ylab="Log-cpm")
```

## Unsupervised clustering of samples
The MDS plot is used to check the similarities and dissimilarities in a sample. The differences can show you the expected results of the differential expression analysis. The "plotMDS" function allows us to create the graph.
```{r}
library(limma)
library(Glimma)
lcpm <- cpm(x, log=TRUE)
par(mfrow=c(1,2))
group
levels(group) <-  brewer.pal(nlevels(group), "Set1")
col.group <- as.character(group)
col.group <- c("purple","orange")[group]
plotMDS(lcpm, labels=group, col=col.group)
title(main="A. Sample groups")
```

```{r}
glMDSPlot(lcpm,groups = group)
```

## Differential Expression Analysis
## Create a design matrix and contrasts
This is the differential expression analysis. First, I create a design matrix with both groups of data.
```{r}
design <- model.matrix(~0+group)
colnames(design) <- gsub("group", "", colnames(design))
design
```

This is the contrasts for the two groups created and stored in a matrix.
```{r}
contr.matrix <- makeContrasts(
   BlackVsWhite = black_male-white_male, 
   levels = colnames(design))
contr.matrix
```

This is the voom plot.
```{r}
par(mfrow=c(1,2))
v <- voom(x, design, plot=TRUE)
v
```

This is our final mean-variance trend.
```{r}
vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts=contr.matrix)
efit <- eBayes(vfit)
plotSA(efit, main="Final model: Mean-variance trend")
```

T-statistics allows for a more stringent definition of significance to be applied to the data. Using the "summary and decide test function", we are looking at which genes are down-regulated and up-regulated.
```{r}
summary(decideTests(efit))
```
```{r}
tfit <- treat(vfit, lfc=0.3)
dt <- decideTests(tfit)
summary(dt)
```
```{r}
de.common <- which(dt[,1]!=0)
length(de.common)
```
```{r}
head(tfit$genes$SYMBOL[de.common], n=20)
```
```{r}
vennDiagram(dt[,1], circle.col=c("turquoise", "salmon"))
```
```{r}
write.fit(tfit, dt, file="results.txt")
```

I use the "toptreat" function. I list the genes that are most differentially expressed.
```{r}
Black.vs.White <- topTreat(tfit, coef=1, n=Inf)
head(Black.vs.White)
```

I use "plotMD" function to summarize the genes. The log fold change is plot against the log counts per million.
```{r}
plotMD(tfit, column=1, status=dt[,1], main=colnames(tfit)[1], 
       xlim=c(-8,13))
```

The same, we can use glimma to make this interactive.
```{r}
glMDPlot(tfit, coef=1, status=dt, main=colnames(tfit)[1],
         side.main="ENSEMBL", counts=lcpm, groups=group, launch=TRUE)
```

Finally, I want to plot the differential expression as a heatmap. Heatmap.plus is a library that allows for the creation of heatmaps for larger datasets.
```{r}
library(gplots)
library(heatmap.plus)
Black.vs.White.topgenes <- Black.vs.White$ENSEMBL[1:40]
i <- which(v$genes$ENSEMBL %in% Black.vs.White.topgenes)
mycol <- colorpanel(1000,"blue","white","red")
par(cex.main=0.8,mar=c(1,1,1,1))
heatmap.plus(lcpm[i,],  scale="row", labRow=v$genes$SYMBOL[i], labCol=group, col=mycol, cexRow=1,cexCol=0.2, margins = c(8,6), main="HeatMap")
```

In this heatmap, red regions are samples with high expression, and blue regions are samples with relatively low expression.

## Gene set testing with camera
I use the camera method to analysize with some gene set testing. The gene set is from the Broad Institute’s Molecular Signatures Database (MSigDB). You can download the Rdata from this link. <http://bioinf.wehi.edu.au/software/MSigDB/> After you download, make sure to move it to your working directory.
```{r}
load("human_c2_v5p2.rdata")
idx <- ids2indices(Hs.c2, rownames(v$genes))
cam.BlackVsWhite <- camera(v,idx,design,contrast=contr.matrix[,1])
head(cam.BlackVsWhite, 7)
```

This barcode plot shows the relative enrichment of black male versus white male in clear cell renal cell carcinoma.
```{r}
barcodeplot(efit$t[,1], index=idx$LIM_MAMMARY_LUMINAL_MATURE_UP, 
            index2=idx$LIM_MAMMARY_LUMINAL_MATURE_DN, main="BlackVsWhite")
```



